Concurrent Programming with CXL

### 2020-06-24 book/paper.tex
Worked on monitor chapter

On the left is a Hoare monitor.  When a process leaves the monitor
(let's call that process $p$), it must be in the bathroom.
It can choose to either let in a process
in one of the bedrooms or a process in the hall, assuming there are any.
To make the choice, $p$ checks for each bedroom if its condition holds
and if there are processes in the bedroom.
If there are such bedrooms, $p$ selects one and lets that process into
the bathroom.  Let's call that process $q$.  Importantly, when $q$ enters
the bathroom, the condition that $q$ was waiting for still holds.

@@

Mesa monitors provide another important option:
when process $p$ leaves the bathroom, it can choose to let any number of
processes in the bedrooms into the hall.  It can do so by calling
\texttt{notify} repeatedly, each time letting one process in one of the
bedrooms into the hall.  Using \texttt{notify} on an empty
bedroom is considered a no-op in Mesa.  (Hoare monitors do not allow
using \texttt{signal} on an empty bathroom.) Mesa also has a function called
\texttt{notifyAll}
\index{notifyAll}
(aka \texttt{broadcast)})
\index{broadcast}
that, when applied to
a bedroom, let's all processes in that bedroom into the hall.

While Hoare monitors are conceptually simpler, the so-called
``Mesa monitor semantics'' or ``Mesa condition variable'' semantics
have become more popular, adopted by all major
programming languages.

That said, few programming languages provide full support for monitors.
In Java, each object has a hidden lock \emph{and} a hidden condition variable
associated with it.
Methods declared with the \texttt{synchronized} keyword automatically
obtain the lock.  Java objects also support \texttt{wait}, \texttt{notify},
and \texttt{notifyAll}.  Java also supports explicit allocations of locks
and condition variables.
In Python, locks and condition variables must be explicitly declared.
The \texttt{with} statement makes it easy to acquire and release a lock
for a section of code.
In C and C++, support for locks and condition variables is through libraries.

@@

The \texttt{synch} module supports Mesa condition variables.  However, like C, Java, and Python,
it does not have built-in language support for them but instead associates condition variables
with explicit locks.
Figure~\ref{fig:cv} shows the \texttt{synch} interface to condition variables.
\texttt{Condition(lk)} creates a new condition variable associated with the given pointer
to a lock \texttt{lk}.
The condition variable itself is represented by a dictionary containing the pointer to the lock
and a bag of name tags of processes waiting on the condition variable.
(The \texttt{synchS} library instead has a list of contexts.)

@@

To illustrate how Mesa condition variables are used in practice, we demonstrate, once again,
using an implementation of reader/writer locks.
Figure~\ref{fig:RWcv} shows the code.  \texttt{rwlock} is the global lock or mutex for the
reader/writer lock.  There are two condition variables: readers wait on \texttt{rcond} and
writers wait on \texttt{wcond}.  The implementation also keeps track of the number of
readers and writers in the critical section.

When inspecting the code, one important feature to notice is that \texttt{wait} is always
invoked within a \texttt{while} loop that checks for the condition that the process is
waiting for.  We explained why in Chapter~\ref{ch:monitors}: when a process resumes after
being notified, it is not guaranteed that the condition it was waiting for holds, even
if it did at the time of invoking \texttt{notify} or \texttt{notifyAll}.
It is \emph{imperative} that you always have a \texttt{while} loop around any invocation
of \texttt{wait}.
Many implementation of condition variables depend on this, and optimizations of the
code allow so-called ``spurious wakeups,'' where \texttt{wait} may sometimes return even
if it has not been notified.

@@

>>>>

On the left is a Hoare monitor.  When a process leaves the monitor
(let's call that process $p$), it must be in the bathroom.
It can choose to either let in a process
in one of the bedrooms or a process in the hall, assuming there are any.
To make the choice, $p$ checks for each bedroom if its condition holds
and if there are processes in the bedroom.
If there are such bedrooms, $p$ selects one and one of its processes into
the bathroom.  Let's call that process $q$.  Importantly, when $q$ enters
the bathroom, the condition that $q$ was waiting for still holds.

@@

Mesa monitors provide another important option:
when process $p$ leaves the bathroom, it can choose to let any number of
processes in the bedrooms into the hall.  It can do so by calling
\texttt{notify} repeatedly, each time letting one process in one of the
bedrooms into the hall.  Using \texttt{notify} on an empty
bedroom is considered a no-op in Mesa.  (Hoare monitors do not allow
using \texttt{signal} on an empty bathroom.) Mesa also provides an
operation called \texttt{notifyAll}
\index{notifyAll}
(aka \texttt{broadcast)})
\index{broadcast}
that, when applied to
a bedroom, lets all processes in that bedroom into the hall.
This is not possible with Hoare monitors: when needing to wake up a set
of processes, only one can be signaled at a time.  Each of those processes
then has to signal the next until all processes have been signaled.

While Hoare monitors are conceptually simpler, the so-called
``Mesa monitor semantics'' or ``Mesa condition variable'' semantics
have become more popular, adopted by all major
programming languages.
That said, few programming languages provide full support for monitors.
In Java, each object has a hidden lock \emph{and} a hidden condition variable
associated with it.
Methods declared with the \texttt{synchronized} keyword automatically
obtain the lock.  Java objects also support \texttt{wait}, \texttt{notify},
and \texttt{notifyAll}.  Java also supports explicit allocations of locks
and condition variables.
In Python, locks and condition variables must be explicitly declared.
The \texttt{with} statement makes it easy to acquire and release a lock
for a section of code.
In C and C++, support for locks and condition variables is through libraries.

@@

The CXL \texttt{synch} module supports Mesa condition variables.
However, like C, Java, and Python,
it does not have built-in language support for them but instead associates condition variables
with explicit locks.
Figure~\ref{fig:cv} shows the implementation of
condition variables in the \texttt{synch} module.
\texttt{Condition(lk)} creates a new condition variable associated with the given pointer
to a lock \texttt{lk}.
The condition variable itself is represented by a dictionary containing the pointer to the lock
and a bag of name tags of processes waiting on the condition variable.
(The \texttt{synchS} library instead uses a list of contexts.)

@@

To illustrate how Mesa condition variables are used in practice, we demonstrate using,
once again, an implementation of reader/writer locks.
Figure~\ref{fig:RWcv} shows the code.  \texttt{rwlock} is the global lock or mutex for the
reader/writer lock.  There are two condition variables: readers wait on \texttt{rcond} and
writers wait on \texttt{wcond}.  The implementation also keeps track of the number of
readers and writers in the critical section.

When inspecting the code, one important feature to notice is that \texttt{wait} is always
invoked within a \texttt{while} loop that checks for the condition that the process is
waiting for.  We explained why above: when a process resumes after
being notified, it is not guaranteed that the condition it was waiting for holds, even
if it did at the time of invoking \texttt{notify} or \texttt{notifyAll}.
It is \emph{imperative} that there is always a \texttt{while} loop around any invocation
of \texttt{wait} containing the negation of the condition that the process is waiting for.
Many implementation of condition variables depend on this, and optimized implementations
of condition variables often allow so-called ``spurious wakeups,'' where \texttt{wait}
may sometimes return even if the conditon variable has not been notified.

@@

Much of the complexity of programming with Mesa condition variables is in figuring out
when to invoke \texttt{notify} and when to invoke \texttt{notifyAll}.  As a rule of
thumb: be conservative---it is better to wake up too many processes than too few.
Waking up too many processes may lead to some inefficiency, but waking up to few
may cause the application to get stuck.  CXL can be particularly helpful here, as
it examines every corner case.  You can safely try to replace each \texttt{notifyAll}
with \texttt{notify} and see if every possible execution of the the application still
terminates.
